---
title: 'BS31005: QTL Mapping and GWAS Workshop'
author: "James Abbott"
date: "11/11/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 100
---

# BS31005: QTL Mapping and GWAS Workshop

Run the setup code chunk below to initialise the R session and load the required libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("farver", quietly = TRUE)) {
    install.packages("farver")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
    install.packages("ggplot2")
}
if (!requireNamespace("qtl", quietly = TRUE)) {
    install.packages("qtl")
}

library(ggplot2)
library(qtl)

data('hyper')

```

## Provided Data

The data for this exercise is provided in a number of CSV files:

-   `heights.csv`: The heights of 500 people, as a single column

-   `bristle_length.csv`: *Drosophila* flies have bristles on their abdomen, which were measured in
    the a number of mothers, and all their offspring. This CSV file contains two columns - the first
    is the average bristle length of the mother, and the second is the average bristle length of all
    her offspring.

-   `pollen_shed_expt.csv`: This contains the results of an investigation into pollen shedding in
    maize, where the number of days between planting and pollen being shed investigated in 10 inbred
    maize lines, with each line grown in three different environments

## Data Loading

Load each of the provided CSV files into a data frame using your preferred method - you may locate
the files in the file browser, click the file name and select `Import Dataset…`, or select 'Import
Dataset' from the `file` menu, or for the really adventurous, try using the read_csv function
directly i.e.

```{r}
heights<-read.csv('heights.csv')
```

If you use the 'read.csv' function, name the data frames with the same name as the CSV files, jut
without the `.csv` suffix, such as in the above example using `heights`.

In the case of `heights`, this single column of data is more useful held as a list rather than a
data-frame of one column, so we can extract this column into a list:

```{r}
heights<-heights[[1]]
```

## Working through this document

You will find a number of empty or partially completed code cells throughout this RMarkdown
document. As you work through the document it will ask you to carry out certain tasks, which are
bullet-pointed text in bold. Enter the code for each of these in the code chunk beneath the relevant
set of instructions, then click the green 'play' triangle icon in the top right corner of the code
chunk.

## Summary Statistics

The statistical methods discussed in the lecture are needed for these examples. Although the
equations look complex, they are actually relatively straightforward to use since R will do the hard
work for us.

We will just be using the `mean()`, `rowMeans()`, `colMeans()`, `var()`, `cov()` and `cor()`
functions. The first section of the workshop will either introduce you to these, or be a bit of a
reminder if you are already familiar with them.

### Mean

The mean of a set of values is formally expressed as$$\bar{X}=\frac{1}{n}\sum_{i=1}^{n} X_i$$

which translates as the more human readable

$$\bar{X}=\frac{X_1 + X_2 ...X_n}{n}$$

In R, we just need to call the `mean()` function and pass it the data we want to find the mean of as
an argument, and the calculation is done for us.

-   **Call the `mean()`** **function and pass it the `heights` list**

```{r mean}

```

If we are working with data in a data frame, We can find the mean of the rows with `rowMeans` and
the mean of the columns with `colMeans`.

-   Call the `rowMeans()` function and pass it the `bristle_length` data frame. This will return the
    results as a list.

```{r df_rowmeans}

```

-   **Similarly, call `colMeans()` and pass it the `bristle_length` data frame.** This time the
    results will be returned as a named list

```{r df_colmeans}

```

### Variance

Variance, a measure of how far values are spread from the mean, is calculated as:
$$V_x = \frac{1}{n}\sum_{i}(X_i-\bar{X})^2$$ Similarly, R can do the hard work for use with the
`var()` function.

-   **Call the `var()` function, passing it the `heights` list.**

```{r variance}

```

### Covariance

Covariance is a measure of the joint variability of two variables.
$$COV_x,_y = \frac{1}{n}\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})$$ Since this is a comparison between two
variables, we need to provide it with two sets of variables. In our case we can do this with the
`bristle_length` data frame. If we pass a data frame to the `cov()` function it will determine the
covariance between all possible combinations of the columns of the data frame, and return the
results as a matrix.

-   **Call the `cov()` function and pass it the bristle_length data frame.**

```{r covariance}

```

We can also obtain the covariance of a single combination of variable by providing the `cov()`
function with two vectors, or columns of data from the data frame (i.e. `bristle_length$mother`).

-   Call the `cov()` function, passing it the `mother` and `offspring_mean` columns as two separate
    arguments i.e. `cov(column1, column2)`

```{r covariance_of_vectors}

```

### Broad-Sense Heritability

This exercise is based on the example in Introduction to Genetic Analysis (Griffiths, Doebley,
Peichel and Wassarman. 12th Edition p.653)

The data in the `pollen_shed_expt.csv` file is a data frame containing the results of an experiment
to investigate pollen shedding time in Maize. The number of days between planting and pollen being
shed was recorded for 10 inbred maize lines, in 3 separate environments. This data allows us to look
at the effect of genetic variation between lines, and also environmental variation between the three
different environments.

Broad-sense heritability (representing the proportion of phenotypic variation resulting from genetic
factors relative to the total phenotypic variation) can be determined using the equation

$$
H_2 = \frac{V_g}{V_x}
$$

where $V_g$ represents the phenotypic variance due to genetic factors, and $V_x$ represents the
total phenotypic variance. $V_x$ can itself be broken down into the contributions from $V_g$ and
also environmental factors, represented by $V_e$, thus

$$
H_2 = \frac{V_g}{V_g + V_e}
$$

We therefore need to determine $V_g$ and $V_e$ for our data. Using the `pollen_shed_expt1` data,

Firstly, we can use the `colMeans()` and `rowMeans()` functions to find the means of the rows and
columns in our data frame, which represent the mean value for the environments and lines
respectively.

-   **Complete the code cell below to capture the environment means and line means**

```{r broad-sense2}
environment_means<-
line_means<-
```

Next, we can determine the genetic variance ($V_g$) and environmental mean ($V_e$) from these
vectors of mean values.

-   **Call the `var()` function on each of these lists of mean values, storing the results of the
    function called with `line_means` as `Vg` and the `environment_means` results as `Ve`.**

```{r broad-sense3}
Vg<-
Ve<-
```

Since $V_x = V_g + V_e$ and have now determined these values, which can find $V_x$ by adding these
values together

-   Complete the following code chunk to determine $V_x$

```{r}
Vx<-
```

Now we can put these values into the broad-sense heriditary equation $$
H_2 = \frac{V_g}{V_x}
$$

-   Complete the following code cell to determine $H_2$ and print out the result. Note that R uses
    the `/` operator for division

```{r broad-sense4}

H2<-
H2

```

Does the environment or genetic effects constitute the main source of variance in this data set?

### Narrow-sense heritability

Narrow-sense heritability is the ratio of the additive genetic variance to the total variance:

$$h^2 = \frac{V_a}{V_x} = \frac{2COV_{p,o}}{V_x}$$

This example uses the results of a study measuring abdominal bristle length in Drosophila. The
length of the bristles of each mother fly were measured, as were those of all the offspring of the
mother, which you should already have loaded from the `bristle_length.csv` file.

```{r narrow-sense1, eval=FALSE, include=FALSE}
View(bristle_length)
```

Plotting this as a scatter plot and fitting a linear regression line suggests there is a
relationship between the bristle length in the mothers and their offspring. This plot is produced
using a more advanced graphing library (ggplot2) than you may be used to - don't worry about the
code here, just the output!

```{r narrow-sense2}
ggplot(bristle_length,aes(x=mother,y=offspring_mean))+
  geom_point()+
  theme_bw()+
  xlab("Mother's bristle length")+
  ylab('Mean offspring bristle length')+
  geom_smooth(method='lm')
```

We can see how tightly linked these two sets of values are by finding the Pearson correlation
coefficient, using the `cor` function.

```{r narrow-sense3}
cor(bristle_length)
```

A correlation coefficient of 1 would indicate the two sets of data were very tightly associated with
each other, while a coefficient of -1 would indicate they have an inverse relationship.
Unsurprisingly (and perhaps reassuringly!) the correlation coefficient of the mother:mother and
offspring_mean:offspring_main combinations are 1.

The key to determining $h^2$ is determining the value of $V_x$ (the total phenotypic variation) and
the covariance between the parents and offspring ($COV_{p,o}$).

-   **Complete the following code chunk to determine the covariance between the bristle length in
    the mothers, and the offspring mean.** Look back at the 'covariance' section if you need a
    reminder on how to do this.

```{r narrow-sense4}
covariance<-
```

Now we need to determine the total phenotypic variance ($V_x$). We can use the R `append` function
to join the two columns of the bristle_lengths dataframe into one vector, then determine the overall
variance within the data.

-   Complete the following code chunk to determine the variance of the `all` list.

```{r narrow-sense6}
all=append(bristle_length$mother,bristle_length$offspring_mean)
Vx<-
```

Next we just need to put everything together. $$h^2=\frac{2COV_{p,o}}{V_x}$$

We have already determined $COV_{p,o}$ which should be stored in the `covariance` variable, so we
just need to multiply this by 2, and divide it by $V_x$

-   Complete the following code chunk to determine $h^2$, and print the result

```{r narrow-sense7}
h2<-
h2
```

What does this tell us about the inheritance of the bristle length phenotype?

### Predicting Phenotypes

We can predict the phenotype of the offspring of a cross if we know the narrow-sense heritability
($h^{2}$) with the breeder's equation:

$$\hat{x}_{o} = h^{2}(\frac{x^{'}+x^{''}}{2})$$

or

$$\hat{x}_o = h^2\bar{x}_p$$

where $\hat{x}_{o}$ represents the offspring's phenotypic deviation from the mean, and $x^{'}$ and
$x^{''}$ are the parent's additive deviations and $\bar{x}_p$ represents the mean parental deviation
from the mean.

This is also sometimes shown in Lush's original form:

$$R=h^{2}S$$

where $R$ represents the response, and $S$ is the selection differential.

### Wheat Yield Exercise

Wheat has an average yield of 3.4 tonnes per hectare.

We cross two varieties, which have yields of 3.1 and 3.8 tonnes per hectare respectively. The
narrow-sense heritability ($h^{2}$) of the trait is 0.63 What do we expect the yield of the
offspring of the cross to be?

-   **Run the below code chunk which will store these values in the mean_yield, yield1, yield2 and
    h2 variables.**

```{r yield1}
mean_yield<-3.4
yield1<-3.1
yield2<-3.8
h2<-0.63
```

First, find the deviance of the parental phenotypes from the mean. Deviation of the sample from the
mean can be found with $$x = X-\overline{x}$$

where $X$ is the yield for the variety, and $\bar{x}$ is the mean yield.

-   **Compete the code chunk below to determine the variance of each parent's yield from the mean**

```{r yield2}
x1<-
x2<-
```

We can now determine the parental mean deviation ($\bar{x}_{p}$) by passing a list of the parental
deviations i.e. `c(x1,x2)` to the `mean()` function.

-   **Complete the code chunk below to determine the mean of x1 and x2 and store it as `mean_xp`**

```{r yield3}
mean_xp<-
```

Now we have determined $\overline{x}_{p}$, and we know the proportion of additive genetic variation
(the narrow-sense heritability $h^{2}$) we can put these into the Breeders equation
$$\hat{x}_o = h^2\bar{x}_p$$ which will give us the offspring's phenotypic deviation from the mean
($\hat{x}_{o}$)

-   **Complete the code chunk below to determine** $\hat{x_o}$

```{r yield4}
xo<-
```

Finally, since we now know the expected deviation of the offsprings yield from the mean yield, we
can just add these together to see what the expected yield of the results of this cross would be

```{r yield5}
predicted_yield<-xo + mean_yield
cat('Predicted yield', predicted_yield, '\n')
```

## QTL Mapping: Hypertension in mice

### Background

This exercise carries out a QTL analysis of a dataset investigating hypertension in mice. The
purpose of the exercise is to walk you through the process of QTL mapping using the `qtl` R library
which does all the hard work for us. You won't be expected to write any R code for this - it is all
provided for you, so you just need to run each code block as you work through the exercise.

The dataset is one of the example sets available as part of the `qtl` library which was published in
[Sugiyama et al.
(2001)](https://www.sciencedirect.com/science/article/pii/S0888754300964012?via%3Dihub){.uri}.

This study aimed to identify loci associated with salt-induced hypertension in mice. Backcrosses
were carried out between salt-sensitive and salt-insensitive inbred strains. The drinking water
provided to the mice contained 1% NaCl and after 14 days, blood pressure measurements were taken
every day for 5 days, and averaged.The mice were then genotyped at a total of 174 markers.

### QTL Data

The raw data used in this exercise is available as a csv file in the module materials: hyper.csv.
You can open this in Excel to see what the data looks like, but a small section is show below.

| bp    | sex  | D1Mit296 | D1Mit123 | D1Mit156 |
|-------|------|----------|----------|----------|
|       |      | 1        | 1        | 1        |
|       |      | 3.3      | 19.7     | 32.8     |
| 109.6 | male | BA       | BA       | BA       |
| 109.8 | male | BA       | BA       | BA       |
| 110.1 | male | BB       | BB       | BB       |

The first two columns contain the phenotypic data - in this case, the mean of the recorded blood
pressures readings for the individual mouse, and it's sex.

The remaining columns contain the genotyping data. Each column has the marker name in the first row,
followed by the chromosome it is located on, and then it's position on the chromosome in cM. For
each of the remaining rows, the genotype of the marker in each mouse is expressed as 'AA', 'AB',
'BA' or 'BB'.

The data is already available in your R session as an object named `hyper`. The `summary` R command
will give us some details about the data.

```{r}
summary(hyper)
```

The `plot` function which will provide an easy visualistion of the data for us to give us a feel for
what it looks like.

```{r}
plot(hyper)
```

This can be hard to make out with overlapping labels, but there are three icons at the top right of
the plot. Clicking the left-hand one of these will open the plot in a separate window then it can be
made full screen.

The first plot indicates where there are missing genotypes in the data, and is pretty impossible to
interpret unless made very large. Second (top right) is the genetic map, indicating the marker
locations on the chromosomes. This study dates from before the mouse genome was available, so marker
density is a lot lower than it would be in a modern study. A lower marker density means the trait
will be mapped less precisely. The lower two plots are showing the recorded phenotypes, firstly
blood pressure (bp), which appears to be normally distributed. The second indicates the sex of the
mice in the study. These are all male, as a result of the experimental design, since male mice of
the salt-sensitive strain show a stronger effect than females. This also makes our lives easier,
since in a mixed population of males and females we need to account for the X and Y chromosomes
differently.

We can inspect the raw phenotype data by viewing the \`pheno\` element of the \`hyper\` object. This
is a data frame i.e. a table, with named columns. The \`View\` command opens this in a new tab in
RStudio for us to look at.

```{r eval=FALSE, include=TRUE}
View(hyper$pheno)
```

We can drop any markers which have no genotype data to clean up the dataset a bit. The
`drop.nullmarkers()` function of the `qtl` package returns a modified version of the 'hyper' object,
so we can store this back into the 'hyper' variable.

```{r}
hyper<-drop.nullmarkers(hyper)
summary(hyper)
```

Errors in genotyping are not uncommon. The `calc.genoprob()` function of the `qtl` package uses a
statistical model called a Hidden Markov Model (HMM) to determine the probabilities of the true
genotypes given the observed data. This function again stores it's outputs in the object, so we
reassign this back to 'hyper' again. The 'step' argument to `calc.genoprob()` indicates the maximum
genetic distance (in centimorgans) between which probabilities are determined, while error.prob
defines the assumed error rate in the genotyping.

```{r}
hyper<-calc.genoprob(hyper, step=1, error.prob=0.01)
```

We can now carry out a scan of the genome to identify QTLs, using the `scanone()` function, and
assign the returned data to the variable `results`.

```{r}
results<-scanone(hyper)
```

Remember from the lecture that the LOD score is the log of the ratio of the probability that there
is a QTL present at the locus to the probability that there is not a QTL present at the locus:

$$LOD = log_{10}\left(\frac{Prob(data|QTL)}{Prob(data|noQTL)}\right)$$

As a log10 value, a LOD score of 2 equates to a ration of 1:100, while LOD 3 = 1:1000 and so on.

We can get an overview of the outputs by passing the results from the scan to the `summary()`
function. This reports the marker with the highest LOD score on each chromosome in the first column,
followed by the chromosome, position (in cM) and LOD score.

```{r results='markup'}
summary(results)
```

You can see the full contents of the results with the \`View\` command, which will display same
details for every marker on the genome:

```{r eval=FALSE, include=TRUE}
View(results)
```

The \`plot\` command can also produce a plot of the genome scan results for us, with the chromosomes
indicated on the \$x\$ axis, and the LOD score on the \$y\$.

```{r}
plot(results, bandcol="gray95")
```

### Permutation Testing

This shows us that there are a number of peaks in LOD score, which may indicate QTLs. The
`scanone()` function also includes the ability to run a permutation test. A permutation test measure
every possible rearrangement of the observed data and assesses it statistically to determine the
'background noise' in the data. This enables the significance threshold of our data to be determined
i.e. the LOD score above which results are not likely to have occurred by chance. The normal cutoff
used to establish significance is 5% ($p<0.05$), which means that for every 100 observations above
the LOD threshold, 5 of these would be expected to have occurred by chance.

There is a nice introduction to permutation tests involving washing alpacas
[here](https://www.jwilber.me/permutationtest/){.uri} if you are interested in learning more.

To run the permutation test, we just need to use the `n.perm` argument to tell `scanone()` how many
permutations to run. 1000 would be ideal, but this is pretty slow so we'll just go for 100 - this
will take a minute or so to run.

```{r results='hide'}
perms<-scanone(hyper, n.perm=100)
```

If we use the \`plot\` function to visualise the results of the permutation test we will see the LOD
scores on the $x$ axis, and the frequency on the $y$ axis. The majority of the data fits into a
normal distribution, with a tail on the right-hand side with a LOD score above three. Values which
fall outside the normal distribution data are indicative of where the LOD cut-off is going to be,
which looks to be around 3.

```{r include=TRUE}
plot(perms)
```

We can find the exact value of the LOD threshold by running the `summary` function, using the
`alpha` argument to define the p-value cut-off to use.

```{r include=TRUE}
summary(perms,alpha=0.05)
```

This shows that the actual significance threshold for a p-value of 0.05 is a LOD score somewhere
around 2.6 to 2.9 - this value will vary slightly since we ran a limited number of permutations -
running a higher number of permutations would produce a more reliable result.

The `summary()` function can also use the results of the permutation test to select the markers with
the highest LOD score on each chromosome which exceed the specified significance threshold. We pass
the `results` object to the function, along with the results of the permutation test (`perms`), the
significance threshold to use (`alpha`) and also request that it reports the p-values of the
selected markers.

```{r include=TRUE}
summary(results,
  perms=perms,
  alpha=0.05,
  pvalues=TRUE)
```

This gives us two markers which meet our significance threshold and seem to be associated with the
trait. The authors of the original paper used some rather more sophisticated techniques and were
able to identify 6 loci associated with the trait, but four of these are not statistically
significant using the approach we have used here.

We can now take a look at where out QTLs are on the genetic map. First we have to construct a `qtl`
object containing the locations of our loci. We pass this the original dataset (`hyper`) which
includes the genetic map data. and pass it a list of the chromosomes (`chr`) carrying the loci, and
there positions (`pos`). The `what` argument allows you to select what to include in the qtl
object - we want to use `prob` to indicate we want the probabilities including.

```{r}
qtl<-makeqtl(hyper,
  chr=c('1', '4'),
  pos=c(48.3, 29.5),
  what='prob')
```

We will create separate plots for each chromosome since they can get quite 'busy' with overlapping
labels. Just as with the previous command we pass a list of chromosomes to plot as \`chr\`, and I've
also set show.marker.names to TRUE to label the markers. If you want a clearer picture you can
remove the marker names by setting \`show.marker.names=FALSE\` instead.

```{r}
plot(qtl,show.marker.names=TRUE, chr=c(1))#, justdots=TRUE)
plot(qtl,show.marker.names=TRUE, chr=c(4))#, justdots=TRUE)
```

We now have identified two loci potentially linked with our trait, which can be used in future work
to try to narrow down the genes involved.
